---
title: "<font style='font-size:1em;'>🗓️ Week 2<br/> Fitting Lines </font>"
author: Dr. [George Melios](#)
institute: '[London School of Economics and Political Science](#)'
date: 09 October 2024
date-meta: 09 October 2024
date-format: "DD MMM YYYY"
toc: true
toc-depth: 1
toc-title: "What we will cover today:"
center-title-slide: false
from: markdown+emoji
format:
  revealjs: 
    fig-responsive: true
    theme: simple
    slide-number: true
    mouse-wheel: false
    preview-links: auto
    logo: /figures/logos/MY_INSTITUTION.png
    css: /css/styles_slides.css
    footer: 'PB4A7- Quantitative Applications for Behavioural Science'
---


```{r setup, include=FALSE}
# options(htmltools.dir.version = FALSE) 
# knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
# #knitr::opts_chunk$set(echo = FALSE, dev.args = list(png = list(type = "cairo")), warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(gghighlight)
library(jtools)
library(Cairo)
library(gganimate)
library(ggdag)
library(jtools)
library(purrr)
library(Cairo)
library(magick)
library(fixest)
library(directlabels)
library(ggrepel)
theme_metro <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16),
        axis.title.x = element_text(hjust = 1),
        axis.title.y = element_text(hjust = 1, angle = 0))
}
theme_metro_regtitle <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
```


# Identification Error

- What is identification error?
- *Identification* is how you link the *result you see* with the *conclusion you draw from it*
- For example, say you observe that kids who play video games are more aggressive in everyday life (result), and you conclude from that result that video games make kids more aggressive (conclusion)

---

## Identification Error

- But the same "result" could give us a different conclusion, like *kids who are aggressive play more video games* or people notice aggression more when kids play video games, then 
- *we have made an identification error - our result was not *identified!*
- Identification error is when your result in the data doesn't actually have a clear theory ("why" or "because")
- For example, if you observe that people tend to wear more shorts on days they eat more ice cream, and you conclude that eating ice cream makes you put on shorts

---

## Data Generating Process

- To avoid this, we think about the *data generating process*
- What is a data generating process?
- The data generating process is the *true set of laws* that determine where our data comes from
- For example, if you hold a rock and drop it, it falls to the floor
- What is the data we observe? (Hold the rock & Rock is up) and (Let go & Rock is down)
- What is the data generating process? Gravity makes the rock fall down when you drop it

---

## Data Generating Process

- Another example is a model of supply and demand
- We observe prices and quantities in a competitive market
- What led to those being the prices and quantities we see?
- The supply and demand model and its equilibrium, we theorize!

---

## Data Generating Process


```{r}
ggplot(tibble(x = c(1:100,1:100), y = c(100:1, 1:100), linetype = c(rep('Theoretical D',100),rep('Theoretical S',100))),
    aes(x=x, y=y, linetype = linetype))+
  geom_line(size = 1.5, color = 'black') + 
  geom_text_repel(aes(label = linetype, x = x + 2), method = 'last.bumpup', size = 16/.pt, nudge_x = 2) + 
  scale_x_continuous(limits = c(1, 120)) +
  geom_point(data = tibble(x = 50.5, y = 50, linetype = 'x'), color = 'red', size = 8) + 
  annotate(geom = 'text', x = 53, y = 50, color = 'red', label = 'Observation', hjust = 0, size = 16/.pt) + 
  labs(x = "Q", y = "P") +
  theme_metro()
```
---

## Data Generating Process

- The prices that we observe come from that theoretical construct
- When we see the prices and quantities moving, according to our theory, it is because the S and D lines are moving
- But we cannot see the S and D lines
- Our goal: use the observations we do see to *infer* what the theoretical model (data generating process) is

--- 


## Data Generating Process

- Harder than it sounds. What inference about S and D can we draw from these observations?

```{r}
ggplot(tibble(x = c(25,35,25,50,60), y = c(35,22,75,54,80)),
       aes(x=x, y=y))+
  geom_point(size = 6, color = 'red') + 
  scale_x_continuous(limits = c(1, 120)) +
  scale_y_continuous(limits = c(0,100)) +
  labs(x = "Q", y = "P") +
  theme_metro()
```

---

## Causality

- A DGP can be described by a series of *equations* describing where the data comes from:

$$ X = \gamma_0 + \gamma_1\varepsilon + \nu $$

$$ Y = \beta_0 + \beta_1X + \varepsilon $$

- This says " $X$ is caused by $\varepsilon$ and $\nu$, and $Y$ is caused by $X$ and $\varepsilon$" - an increase in $X$ *causally increases $Y$* by $\beta_1$
- The goal of econometrics is to be able to estimate what $\beta_1$ is accurately

---

## Causality

- We can also represent this set of relationships as a graph, with arrows telling you what variables cause each other

```{r, dev = 'CairoPNG', fig.width = 6, fig.height = 4}
dag <- dagify(X ~ epsilon + nu,
              Y ~ X + epsilon,
              coords = list(x = c(X = 1, Y = 3, nu = 1, epsilon = 2),
                            y = c(X = 1, Y = 1, nu = 2, epsilon = 2)))
ggdag_classic(dag, size = 5) + 
  theme_dag_blank() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'))
```

---

## Causality

- We do this because most of the relationships we're interested in are causal - we want to know, if we could reach in and *manipulate* $X$, would $Y$ change as a result, and how much?
- Does the minimum wage reduce employment?
- Does quantitative easing avert recessions?
- Does six-sigma improve business performance?
- Does getting an MBA make you a better manager?

---

## Regressions

- In statistics, regression is the practice of *line-fitting*
- We want to *use one variable to predict another*
- Let's say using $X$ to predict $Y$
- We'd refer to $X$ as the "independent variable", and $Y$ as the "dependent variable" (dependent on $X$ that is)
- Regression is the idea that we should characterize the relationship between $X$ and $Y$ as a *line*, and use that line to predict $Y$


---

## $X$ and $Y$

- I have an $X$ value of 2.5 and want to predict what $Y$ will be. What can I do?

```{r, echo = FALSE}
# Initialize tb here
tb <- data.frame(X = c(1, 2, 3, 4, 5), Y = c(2, 4, 1, 3, 5))
ggplot(tb, aes(x = X, y = Y)) + 
  geom_point() + 
  theme_metro() + 
  geom_vline(aes(xintercept = 2.5), linetype = 'dashed', color = 'red') + 
  annotate(geom='label', x = 2.5, y = 13, label = 'X = 2.5', hjust = 0, fill = 'white', size = 16/.pt)
```

---

## $X$ and $Y$

- I can't just say "just predict whatever values of $Y$ we see for $X = 2.5$, because there are multiple of those!
- Plus, what if we want to predict for a value we DON'T have any actual observations of, like $X = 4.3$?

```{r, echo = FALSE}
ggplot(tb, aes(x = X, y = Y)) + 
  geom_point() + 
  theme_metro() + 
  geom_vline(aes(xintercept = 2.5), linetype = 'dashed', color = 'red') + 
  annotate(geom='label', x = 2.5, y = 16, label = 'X = 2.5', hjust = 0, size = 16/.pt, fill = 'white') + 
  geom_vline(aes(xintercept = 4.3), linetype = 'dashed', color = 'red') + 
  annotate(geom='label', x = 4.3, y = 16, label = 'X = 4.3', hjust = 0, size = 16/.pt, fill = 'white') + 
  gghighlight(abs(X - 2.5) < .03)
```

---

## Data is Granular

- If I try to fit *every point*, I'll get a mess that won't really tell me the relationship between $X$ and $Y$
- So, we *simplify* the relationship into a *shape*: a line! The line smooths out those three points around 2.5 and fills in that gap around 4.3

```{r, echo = FALSE}
ggplot(tb, aes(x = X, y = Y)) + 
  geom_point() + 
  theme_metro() + 
  geom_vline(aes(xintercept = 2.5), linetype = 'dashed', color = 'red') + 
  annotate(geom='label', x = 2.5, y = 16, label = 'X = 2.5', hjust = 0, size = 16/.pt, fill = 'white') + 
  geom_vline(aes(xintercept = 4.3), linetype = 'dashed', color = 'red') + 
  annotate(geom='label', x = 4.3, y = 16, label = 'X = 4.3', hjust = 0, size = 16/.pt, fill = 'white') + 
  geom_smooth(method='lm',se=FALSE)
```

---

## Isn't This Worse?

- By adding a line, we are necessarily *simplifying* our presentation of the data. We're tossing out information!
- Our prediction of the *data we have* will be less accurate than if we just make predictions point-by-point
- However, we'll do a better job predicting *other* data (avoiding "overfitting")
- And, since a *shape* is something we can interpret, as opposed to a long list of predictions, which we can't really, the line will do a better job of telling us about the *true underlying relationship*

---

## The Line Does a Few Things:

- We can get a *prediction* of $Y$ for a given value of $X$ (If we follow $X = 2.5$ up to our line we get $Y = 7.6$)
- We see the *relationship*: the line slopes up, telling us that "more $X$ means more $Y$ too!"

```{r, echo = FALSE}
ggplot(tb, aes(x = X, y = Y)) + 
  geom_point() + 
  theme_metro() + 
  geom_segment(aes(x=2.5, xend = 2.5, y = 0, yend = 7.5526), linetype = 'dashed', color = 'red') + 
  annotate(geom='label', x = 2.5, y = 3, label = 'X = 2.5', hjust = 0, size = 16/.pt, fill = 'white') + 
  geom_segment(aes(x= 4.3, xend = 4.3, y = 0, yend = 12.82084), linetype = 'dashed', color = 'red') + 
  annotate(geom='label', x = 4.3, y = 5, label = 'X = 4.3', hjust = 0, size = 16/.pt, fill = 'white') + 
  geom_smooth(method='lm',se=FALSE) + 
  geom_segment(aes(x = 0, xend = 2.5, y = 7.5526, yend = 7.5526), linetype = 'dashed', color = 'red') + 
  annotate(geom = 'label', x = .5, y = 7.5526, label = 'Y = 7.6', vjust = 1, size = 16/.pt, fill = 'white') + 
  geom_segment(aes(x = 0, xend = 4.3, y = 12.82084, yend = 12.82084), linetype = 'dashed', color = 'red') + 
  annotate(geom = 'label', x = .5, y = 12.82084, label = 'Y = 12.8', vjust = 1, size = 16/.pt, fill = 'white')
```


---

## Lines

- That line we get is the *fit* of our model
- A model "fit" means we've taken a *shape* (our line) and picked the one that best fits our data
- All forms of regression do this
- Ordinary least squares specifically uses a *straight line* as its shape
- The resulting line we get can also be written out as an actual line, i.e.

$$ Y = intercept + slope*X $$

---

## Lines

- We can use that line as... a line!
- If we plug in a value of $X$, we get a prediction for $Y$
- Because these $Y$ values are predictions, we'll give them a hat $\hat{Y}$

$$ Y = 3 + 4*X $$

$$ \hat{Y} = 3 + 4*(3.2) $$

$$ \hat{Y} = 15.8 $$

---

## Lines

- We can also use it to explain the relationship
- Whatever the intercept is, that's what we predict for $Y$ when $X = 0$

$$ Y = 3 + 4*X $$

$$ \hat{Y} = 3 + 4*0 $$

$$ \hat{Y} = 3 $$

---

## Lines

- And as $X$ increases, we know how much we expect $Y$ to increase because of the slope

$$ Y = 3 + 4*X $$

$$ \hat{Y} = 3 + 4*3 = 15 $$

$$ \hat{Y} = 3 + 4*4 = 19 $$

- When $X$ increases by $1$, $Y$ increases by the slope (which is $4$ here)

---

## Ordinary Least Squares

- Regression fits a *shape* to the data
- OLS specifically fits a *straight line* to the data described using an $intercept$ and a $slope$
- When we plug an $X$ into the line, we get a prediction for $Y$, which we call $\hat{Y}$
- When $X = 0$, we predict $\hat{Y} = intercept$
- When $X$ increases by $1$, our prediction of $Y$ increases by the $slope$
- If $slope > 0$, $X$ and $Y$ are positively related/correlated

---

## Concept Checks

- How does producing a *line* let us use $X$ to predict $Y$? 
- If our line is $Y = 5 - 2*X$, explain what the $-2$ means in a sentence
- Not all of the points are exactly on the line, meaning some of our predictions will be wrong! Should we be concerned? Why or why not?

---

## How?

- We know that regression fits a line
- But how does it do that exactly?
- It picks the line that produces the *smallest squares*
- Thus, "ordinary least squares"

---

## Predictions and Residuals

- Whenever you make a prediction of any kind, you rarely get it *exactly right*
- The difference between your prediction and the actual data is the *residual*

$$ Y = 3 + 4*X $$

If we have a data point where $X = 4$ and $Y = 18$, then

$$ \hat{Y} = 3 + 4*4 = 19 $$

Then the *residual* is $Y - \hat{Y} = 18 - 19 = -1$.

---

## Predictions and Residuals

So really, our relationship doesn't look like this...

$$ Y = intercept + slope*X $$

Instead, it's...

$$ Y = intercept + slope*X + residual $$

We still use $intercept + slope*X$ to predict $Y$ though, so this is also

$$ Y = \hat{Y} + residual $$

---

## Ordinary Least Squares

- As you'd guess, a good prediction should make the residuals as small as possible
- We want to pick a line to do that
- And in particular, we're going to *square* those residuals, so the really-big residuals count even more. We really don't want to have points that are super far away from the line!
- Then, we pick a line to minimize those squared residuals ("least squares")

---

## Ordinary Least Squares

- Start with our data

```{r}
tb2 <- tibble(X = c(1,3,5,6.5), Y = c(2,6, 3.6, 8)) %>%
  mutate(line1 = 1.5 + X,
         line2 = 2 + .6*X,
         line3 = 1.7324 + .8175*X)
ggplot(tb2, aes(x = X, y = Y)) + 
  geom_point(size = 4) + 
  expand_limits(x = c(0,7), y = c(1,9))+
  theme_metro() + 
  coord_fixed()
```

---

## Ordinary Least Squares

- Let's just pick a line at random, not necessarily from OLS

```{r}
ggplot(tb2, aes(x = X, y = Y)) + 
  geom_point(size = 4) + 
  expand_limits(x = c(0,7), y = c(1,9))+
  theme_metro() + 
  coord_fixed() + 
  geom_line(aes(x = X, y = line1), color = 'blue', size = 1.5)
```

---


## Ordinary Least Squares

- The vertical distance from point to line is the residual

```{r}
ggplot(tb2, aes(x = X, y = Y)) + 
  geom_point(size = 4) + 
  expand_limits(x = c(0,7), y = c(1,9))+
  theme_metro() + 
  coord_fixed() + 
  geom_segment(aes(x = X, xend = X, y = Y, yend = line1), size = 1.5) + 
  geom_line(aes(x = X, y = line1), color = 'blue', size = 1.5)
```

---

## Ordinary Least Squares

- Now square those residuals

```{r}
ggplot(tb2, aes(x = X, y = Y)) + 
  geom_point(size = 4) + 
  expand_limits(x = c(0,7), y = c(1,9))+
  theme_metro() + 
  geom_rect(aes(xmin = X, xmax = X + abs(line1 - Y), ymin = Y, ymax = line1), alpha = .5, fill = 'gray') + 
  geom_line(aes(x = X, y = line1), color = 'blue', size = 1.5) +
  coord_fixed()
```

---

## Ordinary Least Squares

- Can we get the total area in the squares smaller with a different line?

```{r}
ggplot(tb2, aes(x = X, y = Y)) + 
  geom_point(size = 4) + 
  expand_limits(x = c(0,7), y = c(1,9))+
  theme_metro() + 
  geom_rect(aes(xmin = X, xmax = X + abs(line2 - Y), ymin = Y, ymax = line2), alpha = .5, fill = 'gray') + 
  geom_line(aes(x = X, y = line2), color = 'blue', size = 1.5) +
  coord_fixed()
```

---

## Ordinary Least Squares

- Ordinary Least Squares, I can promise you, gets it the smallest

```{r}
ggplot(tb2, aes(x = X, y = Y)) + 
  geom_point(size = 4) + 
  expand_limits(x = c(0,7), y = c(1,9))+
  theme_metro() + 
  geom_rect(aes(xmin = X, xmax = X + abs(line3 - Y), ymin = Y, ymax = line3), alpha = .5, fill = 'gray') + 
  geom_line(aes(x = X, y = line3), color = 'blue', size = 1.5) +
  coord_fixed()
```

--- 

## Ordinary Least Squares

- How does it figure out which line makes the smallest squares?
- There's a mathematical formula for that!
- First, instead of thinking of $intercept$ and $slope$, we reframe the line as having *parameters* we can pick

$$ Y = intercept + slope*X + residual $$

$$ Y = \beta_0 + \beta_1X + \varepsilon $$

---

## Terminology Sidenote

$$ Y = \beta_0 + \beta_1X + \varepsilon $$

- In metrics, Greek letters represent "the truth" - in *the true process by which the data is generated*, a one-unit increase in $X$ is related to a $\beta_1$ increase in $Y$
- When we put a hat on anything, that is our *prediction* or *estimation* of that true thing. $\hat{Y}$ is our prediction of $Y$, and $\hat{\beta_1}$ is our estimate of what we think the true $\beta_1$ is
- Note  "residual" =/= $\varepsilon$ - residuals are what's actually left over from our prediction with real data, but the *error* $\varepsilon$ is the *true* difference between our line and $Y$.
---

## OLS Simulator

- Take a look at this OLS simulator and play around with it: [https://econometricsbysimulation.shinyapps.io/OLS-App/](https://econometricsbysimulation.shinyapps.io/OLS-App/)
- Click "Show Residuals" to turn that on
- Try different data generating processes and standard deviations
- What settings make the residuals small or large? Any guesses why?
- What happens if you take the intercept out? What does that make our line do?
- How close does the line come to the data generating process? Intercept and slope are in the second table below the graph under "Estimate"

--- 

## Concept Checks

- Why might we want to minimize squared residuals rather than just residuals?
- What's the difference between a residual and an error?
- If I have the below OLS-fitted line from a dataset of children:

$$ Height (Inches) = 18 + 2*Age$$

And we have the kids Darryl who is 10 years old and 40 inches tall, and Bijetri who is 9 years old and 37 inches tall, what are each of their: (a) predicted values, (b) residuals, and then what is the sum of their squared residuals?

---

## Recap

- Regression is fitting a *shape* to data to explain the relationship more generally
- Ordinary least squares fits a *straight line* that minimizes the *sum of squared residuals*
- That line has an intercept $\beta_0$ (Our prediction $\hat{Y}$ when $X = 0$) and a slope $\beta_1$ (how much higher we predict $Y$ will be when we look at an $X$ one unit higher)
- The *residual* is the difference between our prediction $\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1X$ and the actual number $Y$

$$ Y = \beta_0 + \beta_1 X + \varepsilon $$
$$ \hat{Y} = \hat{\beta}_0 + \hat{\beta}_1X $$

---
