---
subtitle: "PB4A7- Quantitative Applications for Behavioural Science"
title: "<font style='font-size:1em;'>🗓️ Week 7<br/> Interactions & beyond ols </font>"
author: Dr. [George Melios](#)
institute: '[London School of Economics and Political Science](#)'
date: 13 November 2024
date-meta: 13 November 2024
date-format: "DD MMM YYYY"
toc: true
toc-depth: 1
toc-title: "What we will cover today:"
center-title-slide: false
from: markdown+emoji
format:
  revealjs: 
    fig-responsive: true
    theme: simple
    slide-number: true
    mouse-wheel: false
    preview-links: auto
    logo: /figures/logos/MY_INSTITUTION.png
    css: /css/styles_slides.css
    footer: 'PB4A7- Quantitative Applications for Behavioural Science'
---
<!-- ---
title: "Binary Variables and Functional Form"
subtitle: "i.e. what you actually do most of the time"
date: "Updated `r Sys.Date()`"
output:
  xaringan::moon_reader:
    self_contained: TRUE
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    # Run xaringan::summon_remark() for this
    #chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
--- -->


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE) 
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(dagitty)
library(ggdag)
library(fixest)
library(scales)
library(Cairo)
library(estimatr)
library(ggthemes)
library(directlabels)
library(marginaleffects)
theme_metro <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16),
        axis.title.x = element_text(hjust = 1),
        axis.title.y = element_text(hjust = 1, angle = 0))
}
theme_void_metro <- function(x) {
  theme_void() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
theme_metro_regtitle <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
```

```{css, echo=FALSE}
pre {
  max-height: 350px;
  overflow-y: auto;
}

pre[class] {
  max-height: 100px;
}
```

## Interactions

- For both polynomials and logarithms, the effect of a one-unit change in $X$ differs depending on its current value (for logarithms, a 1-unit change in $X$ is different percentage changes in $X$ depending on current value)
- But why stop there? Maybe the effect of $X$ differs depending on the current value of *other variables!* - Enter interaction terms!

$$ Y = \beta_0 + \beta_1X + \beta_2Z + \beta_3X\times Z + \varepsilon $$

- Interaction terms are *a little tough* but also **extremely important**. 

---

## Interactions

Expect to come back to these slides, as you're almost certainly going to use interaction terms in both our assessment and the dissertation


---

## Interactions

- Change in the value of a *control* can shift a regression line up and down
- Using the model $Y = \beta_0 + \beta_1X + \beta_2Z$, estimated as $Y = .01 + 1.2X + .95Z$:

```{r}
set.seed(500)
df <- tibble(X = runif(200), Z = 5*runif(200)) %>%
  mutate(Y = X + Z + rnorm(200))
m <- lm(Y~X+Z, data = df)
mc <- coef(m)[1] %>% unname()
mx <- coef(m)[2] %>% unname()
mz <- coef(m)[3] %>% unname()
ggplot(df, aes(x = X, y = Y)) + 
  geom_point() +
  theme_metro() + 
  scale_x_continuous(limits = c(0,1.1)) +
  geom_line(aes(x = X, y = mc + mx*X + mz), color = 'red', size = 1) + 
  geom_line(aes(x = X, y = mc + mx*X + 3*mz), color = 'red', size = 1) + 
  geom_line(aes(x = X, y = mc + mx*X + 5*mz), color = 'red', size = 1) + 
  annotate(geom = 'text', x = 1.01, y = mc + mx*1.05 + mz, color = 'red', label = 'Prediction\nat Z = 1', hjust = 0, size = 15/.pt) +
  annotate(geom = 'text', x = 1.01, y = mc + mx*1.05 + 3*mz, color = 'red', label = 'Prediction\nat Z = 3', hjust = 0, size = 15/.pt) +
  annotate(geom = 'text', x = 1.01, y = mc + mx*1.05 + 5*mz, color = 'red', label = 'Prediction\nat Z = 5', hjust = 0, size = 15/.pt)

```

---

## Interactions

- But an interaction can both shift the line up and down AND change its slope
- Using the model $Y = \beta_0 + \beta_1X + \beta_2Z + \beta_3X\times Z$, estimated as $Y = .035 + 1.14X + .94Z + 1.02X\times Z$:

```{r}
set.seed(500)
df <- tibble(X = runif(200), Z = 5*runif(200)) %>%
  mutate(Y = X+Z+X*Z + rnorm(200))
m <- lm(Y~X*Z, data = df)
mc <- coef(m)[1] %>% unname()
mx <- coef(m)[2] %>% unname()
mz <- coef(m)[3] %>% unname()
mxz <- coef(m)[4] %>% unname()
ggplot(df, aes(x = X, y = Y)) + 
  geom_point() +
  theme_metro() + 
  scale_x_continuous(limits = c(0,1.2)) +
  geom_line(aes(x = X, y = mc + mx*X + mz + mxz*X), color = 'red', size = 1) + 
  geom_line(aes(x = X, y = mc + mx*X + 3*mz + 3*mxz*X), color = 'red', size = 1) + 
  geom_line(aes(x = X, y = mc + mx*X + 5*mz + 5*mxz*X), color = 'red', size = 1) + 
  annotate(geom = 'text', x = 1.01, y = mc + mx*1.05 + mz + mxz*1.05, color = 'red', label = 'Prediction\nat Z = 1', hjust = 0, size = 15/.pt) +
  annotate(geom = 'text', x = 1.01, y = mc + mx*1.05 + 3*mz  + 3*mxz*1.05, color = 'red', label = 'Prediction\nat Z = 3', hjust = 0, size = 15/.pt) +
  annotate(geom = 'text', x = 1.01, y = mc + mx*1.05 + 5*mz + 5*mxz*1.05, color = 'red', label = 'Prediction\nat Z = 5', hjust = 0, size = 15/.pt)

```

---

## Interactions

- How can we interpret an interaction?
- The idea is that the interaction shows how *the effect of one variable changes as the value of the other changes*
- The derivative helps!

$$ Y = \beta_0 + \beta_1X + \beta_2Z + \beta_3X\times Z $$
$$ \partial Y/\partial X = \beta_1 + \beta_3 Z $$

- The effect of $X$ is $\beta_1$ when $Z = 0$, or $\beta_1 + \beta_3$ when $Z = 1$, or $\beta_1 + 3\beta_3$ if $Z = 3$!

---

## Interactions

- Often we are doing interactions with binary variables to see how an effect differs across groups
- Now, instead of the intercept giving the baseline and the binary coefficient giving the difference, the coefficient on $X$ is the baseline effect of $X$ and the interaction is the difference in the effect of $X$
- The interaction coefficient becomes "the difference in the effect of $X$ between the $Z$ = "No" and $Z$ = "Yes" groups"
- (What if it's continuous? Mathematically the same but the thinking changes - the interaction term is the difference in the effect of $X$ you get when increasing $Z$ by one unit)

---

## Notes on Interactions

- Like with polynomials, the coefficients on their own now have little meaning and must be evaluated alongside each other. $\beta_1$ by itself is just "the effect of $X$ when $Z = 0$", not "the effect of $X$"
- Yes, you *do* almost always want to include both variables in un-interacted form and interacted form. Otherwise the interpretation gets very thorny

---

## Notes on Interactions
- Interaction effects are *poorly powered*. You need a *lot* of data to be able to tell whether an effect is different in two groups. If $N$ observations is adequate power to see if the effect itself is different from zero, you need a sample of roughly $16\times N$ to see if the difference in effects is nonzero. Sixteen times!!
- It's tempting to try interacting your effect with everything to see if it's bigger/smaller/nonzero in some groups, but because it's poorly powered, this is a bad idea! You'll get a lot of false positives

---

---

# OLS and the Dependent Variable

A typical OLS equation looks like:

$$ Y = \beta_0 + \beta_1X + \varepsilon $$

and assumes that the error term, $\varepsilon$, is normal. 

- The normal distribution is continuous and smooth and has infinite range
- And the linear form stretches off to infinity in either direction as $X$ gets small or big
- Both of these imply that the dependent variable, $Y$, is continuous and can take any value (why is that?)!
- If that's not true, then our model will be *misspecified* in some way

---

## Non-Continuous Dependent Variables

When might dependent variables not be continuous and have infinite range?

- Years working at current job (can't be negative)
- Are you self-employed? (Binary)
- Number of children (must be a round number, can't be negative)
- Which brand of soda did you buy? (categorical)
- Did you recover from your disease? (binary)
- How satisfied are you with your purchase on a 1-5 scale? (must be a round number from 1 to 5, and the difference between 1 and 2 isn't necessarily the same as the difference between 2 and 3)

---

## Binary Dependent Variables

- In many cases, such as variables that must be round numbers, or can't be negative, even though there are ways of properly handling these issues, people will *usually* ignore the problem and just use OLS, as long as the data is continuous-ish (i.e. doesn't have a LOT of observations right at 0 next to the impossible negative values, or has lots of different values so the round number smooth out)
- However, the problems of using OLS are a bit worse for binary data, and so they're the most common case in which we do something special to account for it
- Binary dependent variables are also really common! We're often interested in whether a certain outcome happened or didn't (if we want to know if a drug was effective, we are likely asking if you are cured or not!)

So, how can we deal with having a binary dependent variable, and why do they give OLS such problems?

---

## The Linear Probability Model

- First off, let's ignore the completely unexplained warnings I've just given you and do it with OLS anyway, and see what happens
- Running OLS with a binary dependent variable is called the "linear probability model" or LPM

$$ D = \beta_0 + \beta_1X + \varepsilon $$

Throughout these slides, let's use $D$ to refer to a binary variable 

---

## The Linear Probability Model

- In terms of *how we do it*, the interpretation is the exact same as regular OLS, so you can bring in all your intuition
- The only difference is that our interpretation of the dependent variable is now in probability terms
- If $\hat{\beta}_1 = .03$, that means that a one-unit increase in $X$ is associated with a three percentage point increase in the probability that $D = 1$
- (percentage points! Not percentage - an increase from .1 to .13, say, not .1 to .103)

---

## The Linear Probability Model

So what's the problem?

The linear probability model can lead to...

- Terrible predictions
- Incorrect slopes that don't acknowledge the boundaries of the data

---

## Terrible Predictions

- OLS fits a straight line. So if you increase or decrease $X$ enough, eventually you'll predict that the probability of $D = 1$ is bigger than 1, or lower than 0. Impossible!
- We can address part of this by just not trying to predict outside the range of the data, but if $X$ has a lot of variation in it, we might get those impossible predictions even for values in our data. And what do we do with that?
- (Also, because errors tend to be small for certain ranges of $X$ and large for others, we *have* to use heteroskedasticity-robust standard errors)

---

## Terrible Predictins

```{r}
tib <- tibble(X = runif(250)*10) %>%
  mutate(D = ((X - 8)/6 + rnorm(250)*.5 > 0)*1)
ggplot(tib, aes(x = X, y = D)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  theme_metro() + 
  scale_y_continuous(breaks = c(0,1)) + 
  annotate(geom = 'text', x = 6.5, y = .6, color = 'blue', label = 'OLS Prediction')
```

---

## Incorrect Slopes

- Also, OLS requires that the slopes be constant
- (Not necessarily if you use a polynomial or logarithm, but the following critique still applies)
- This is *not what we want* for binary data!
- As the prediction gets really close to 0 or 1, the slope should flatten out to nothing
- If we predict there's a .50 chance of $D = 1$, a one-unit increase in $X$ with $\hat{\beta}_1 = .03$ would increase that to .53
- If we predict there's a .99 chance of $D = 1$, a one-unit increase in $X$ with $\hat{\beta}_1 = .03$ would increase that to 1.02...
- Uh oh! The slope *should* be flatter near the edges. We need the slope to vary along the range of $X$

---

## Incorrect Slopes

- We can see how much the OLS slopes are overstating changes in $D$ as $X$ changes near the edges by comparing an OLS fit to just regular ol' local means, with no shape imposed at all
- We're not forcing the red line to flatten out - it's doing that naturally as the mean can't possibly go any lower! OLS barrels on through though

```{r}
ggplot(tib, aes(x = X, y = D)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + 
  geom_smooth(se = FALSE, color = 'red') +
  theme_metro() + 
  scale_y_continuous(breaks = c(0,1)) + 
  annotate(geom = 'text', x = 5, y = .35, color = 'blue', label = 'OLS Prediction') +
  annotate(geom = 'text', x = 7, y = .25, color = 'red', label = 'Local Means')
```

---

## Linear Probability Model

So what can we make of the LPM?

- Bad if we want to make predictions
- Bad at estimating slope if we're looking near the edges of 0 and 1
- (which means it's especially bad if the average of $D$ is near 0 or 1)

When might we use it anyway?

- It behaves better in small samples than methods estimated by maximum likelihood (which many other methods are)
- If we only care about slopes far away from the boundaries
- If alternate methods (like we're about to go into) put too many other statistical demands on the data (OLS is very "easy" from a computational standpoint)
- If we're using lots of fixed effects (OLS deals with these far more easily than nonlinear methods)
- If our right-hand side is just binary variables (if X has limited range it might not predict out of 0-1!)

---

## Generalized Linear Models

- So LPM has problems. What can we do instead?
- Let's introduce the concept of the *Generalized Linear Model*

Here's an OLS equation:

$$ Y = \beta_0 + \beta_1X + \varepsilon $$

Here's a GLM equation:

$$ E(Y | X) = F(\beta_0 + \beta_1X) $$

Where $F()$ is *some function*. 

---

## Generalized Linear Models

$$ E(D | X) = F(\beta_0 + \beta_1X) $$

- We can call the $\beta_0 + \beta_1X$ part, which is the same as in OLS, the *index function*. It's a linear function of our variable $X$ (plus whatever other controls we have in there), same as before
- But to get our prediction of what $Y$ will be conditional on what $X$ is ( $D|X$ ), we do one additional step of running it through a function $F()$ first. We call this function a *link function* since it links the index function to the outcome
- If $F(z) = z$, then we're basically back to OLS
- But if $F()$ is nonlinear, then we can account for all sorts of nonlinear dependent variables!

So in other words, our prediction of $D$ is still based on the linear *index*, but we run it through some nonlinear function first to get our nonlinear output!

---

## Generalized Linear Models

We can also think of this in terms of the *latent variable* interpretation

$$ D^* = \beta_0 + \beta_1X $$

Where $D^*$ is an unseen "latent" variable that can take any value, just like a regular OLS dependent variable (and roughly the same in concept as our index function)

And we convert that latent variable to a proabability using some function

$$ E(D | X) = F(D^*) $$

and perhaps saying something like "if we estimate $Y^*$ is above the number $c$, then we predict $D = 1$ "


---

## Probit and Logit

- Let's go back to our index-and-function interpretation. What function should we use?
- (many many different options depending on your dependent variable - poisson for count data, log link for nonnegative skewed values, multinomial logit for categorical data...)
- For binary dependent variables the two most common link functions are the probit and logistic links. We often call a regression with a logistic link a "logit regression"

$$ Probit(index) = \Phi(index) $$

where $\Phi()$ is the standard normal cumulative distribution function (i.e. the probability that a random standard normal value is less than or equal to $index$ )

$$ Logistic(index) = \frac{e^{index}}{1+e^{index}} $$

For most purposes it doesn't matter whether you use probit or logit, but logit is getting much more popular recently (due to its common use in data science - it's computationally easier) so we'll focus on that, and just know that pretty much all of this is the same with probit

---

## Logit

- Notice that we can't possibly predict a value outside of 0 or 1, no matter how wild $X$ and our index get
- As $index$ goes to $-\infty$, 

$$ Logistic(index) \rightarrow  \frac{0}{1+0} = 0 $$

- And as $index$ goes to $\infty$,

$$ Logistic(index) \rightarrow  \frac{\infty}{1+\infty } = 1 $$

---

## Logit

- Also notice that, like the local means did, its slope flattens out near the edges

```{r}

ggplot(tib, aes(x = X, y = D)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE, color = 'blue') + 
  geom_smooth(method = 'glm', method.args = list(family = binomial(link = 'logit')), 
              se = FALSE, color = 'red') +
    geom_smooth(method = 'glm', method.args = list(family = binomial(link = 'probit')), 
              se = FALSE, color = 'purple') +
  theme_metro() + 
  scale_y_continuous(breaks = c(0,1)) + 
  annotate(geom = 'text', x = 5, y = .35, color = 'blue', label = 'OLS Prediction') +
  annotate(geom = 'text', x = 7, y = .25, color = 'red', label = 'Logit') +
  annotate(geom = 'text', x = 7, y = .2, color = 'purple', label = 'Probit') 
```

---